Explore the SingleR results

### 2025-07-24 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd
add barplots and PDX section

## Inferences from libraries with lower cell counts

@@

@@

Next, we'll ask whether the percent of unclassified cells (aka, `unknown_singler`) is much higher in these libraries compared to larger libraries.

@@

celltype_recoded_wide_df |>
  dplyr::add_count(library_id, name = "total_cells") |>
  dplyr::inner_join(unknown_singler_df, by = "library_id") |>
  dplyr::select(library_id, total_cells, unknown_cells) |>
  dplyr::distinct() |>
  dplyr::mutate(percent_singler_unclassified = round(unknown_cells/total_cells, 2)) |>
  dplyr::arrange(total_cells)
```

Most cells in the relatively smaller libraries are classified, and the percentage of unclassified cells (when non-0) is fully in line with the larger libraries.

>>>>

## Inferences from smaller vs. larger libraries

@@

### Distirbution of cell types

```{r, fig.width = 10, fig.height = 4}
celltype_recoded_wide_df |>
  dplyr::mutate(
    library_size = ifelse(library_id %in% libraries_low_cells$library_id, "<1000 cells", ">1000 cells")
  ) |>
  ggplot() + 
  aes(x = library_id, fill = singler) + 
  geom_bar(position = "fill") + 
  scale_fill_manual(values = recoded_celltype_pal) + 
  facet_wrap(vars(library_size), scales = "free_x") + 
  scale_y_continuous(expand = c(0,0)) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_blank()
  )
```
Although there are only four libraries with <1000 cells so it is challenging to make definitive conclusions, it does not seem like the overall cell type distributions are very different between smaller and larger libraries.

@@

Next, we'll ask whether the percent of unclassified cells (aka, `unknown_singler`) can be explained by library size.
Below, we plot the relationship between total cells and fraction of cells that SingleR failed to confidently classify.

@@

percent_unclassified_df <- celltype_recoded_wide_df |>
  dplyr::add_count(library_id, name = "total_cells") |>
  dplyr::inner_join(unknown_singler_df, by = "library_id") |>
  dplyr::select(library_id, total_cells, unknown_cells) |>
  dplyr::distinct() |>
  dplyr::mutate(frac_singler_unclassified = unknown_cells/total_cells)

ggplot(percent_unclassified_df) + 
  aes(x = total_cells, y = frac_singler_unclassified) + 
  geom_point() + 
  geom_smooth(method = "lm")
```

There doesn't appear to be a relationship here, and further the percent of unclassified cells are all under 7\% with most below around 4\%.

## Inferences from PDX libraries

In this section, we briefly look at annotation results for libraries from PDX samples, which may show distinct trends.
Here is the breakdown of the number of PDX vs tissue libraries:

```{r}
# This was being developed while this bug was present, so we have to check the type first
# https://github.com/AlexsLemonade/scpcaTools/issues/302
pdx_class <- class(colData(merged_sce)$is_xenograft)
if (pdx_class == "character") {
  pdx_val <- "TRUE"
} else {
  pdx_val <- TRUE
}

pdx_libraries <- colData(merged_sce) |>
  as.data.frame() |>
  dplyr::filter(is_xenograft == pdx_val) |>
  dplyr::pull(library_id) |>
  unique()

celltype_recoded_wide_df <- celltype_recoded_wide_df |>
  dplyr::mutate(sample_type = ifelse(library_id %in% pdx_libraries, "pdx", "tissue"))

celltype_recoded_wide_df |>
  dplyr::select(library_id, sample_type) |>
  dplyr::distinct() |>
  dplyr::count(sample_type)
```

### Distribution of cell types

We'll first look at the cell type distributions for each library between sample types:

```{r, fig.width = 14, fig.height = 6}
ggplot(celltype_recoded_wide_df) + 
  aes(x = library_id, fill = singler) + 
  geom_bar(position = "fill") + 
  scale_fill_manual(values = recoded_celltype_pal) + 
  facet_wrap(vars(sample_type), scales = "free_x") + 
  scale_y_continuous(expand = c(0,0)) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_blank()
  )
```

PDX samples have, not unexpectedly, much less cell type diversity overall with most cells being tumor or similar and immune cells rarely present.
This is consistent with expectations in general for the biology of PDX's.

This section shows a heatmap compaing SingleR annotations to consensus cell types, for PDX samples only.

```{r, fig.width = 8, fig.height = 8}
make_jaccard_heatmap(
  celltype_recoded_wide_df |> dplyr::filter(sample_type == "pdx"),
  "consensus_validation",
  "singler",
  "Consensus validation label",
  "SingleR NBAtlas label"
)
```
This heatmap comparison seems reasonable, again taking into consideration that there are relatively fewer cell types in the PDX samples, and corresponding cell types tend to match up.
